---
title: "How Does a Bike-Share Navigate Speedy Success"
author: "Stephen Nantz"
date: "`r Sys.Date()`"
output: html_document
---

## Cyclistic Case Study

#### Notes:

This case study documents the usage of Cyclistic bikes and how subscription members and casual customers use the service differently.

The data used for this study was made available by Motivate International Inc. under this [license](https://www.divvybikes.com/data-license-agreement). The public data used can be found here: <https://divvy-tripdata.s3.amazonaws.com/index.html>. To keep things simple, a sample of data (2%) from the last 13 months has been included. 

Using 2% of the data provided gives us a confidence level of 99% and a margin of error of 0.1% using this [Sample Size Calculator](https://www.calculator.net/sample-size-calculator.html?type=2&cl2=99&ss2=127742&pc2=2&ps2=6387100&x=53&y=15#findci).

```{r include = FALSE}
library(tidyverse)  
library(janitor)
library(labeling)
library(lubridate)
library(magrittr)
library(readr)
library(ggplot2)
library(dplyr)
options(dplyr.summarise.inform = FALSE)
```

## Cleaning the Data:

The first step in this analysis will be checking the data for errors and removing entries that do not pass scrutiny. A sample size is taken from this cleaned data to keep the file sizes manageable and shareable.

### Data is Cleaned, Shuffled, and Sampled

13 months of data is cleaned, shuffled, and sampled from. We use 2% of the cleaned trip data. Each month is sampled from individually. This ensures a summer month with 600,000 trip entries will supply a sample 12,000 entries (2%), while a winter month with only 50,000 entries will only supply a sample of 1,000 entries (2%).

A 'while' loop is used to prevent massive amounts of data from being kept in memory. A .csv file is written to at the end of each loop.

The file created by this code is the file provided for the report (monthly_data_total.csv). The code in the block below has been provided for reference only.

```{r eval=FALSE}

# 13 files are added to a list
  file_list <- list('202105-divvy-tripdata.csv', '202106-divvy-tripdata.csv', '202107-divvy-tripdata.csv', '202108-divvy-tripdata.csv', '202109-divvy-tripdata.csv', '202110-divvy-tripdata.csv', '202111-divvy-tripdata.csv', '202112-divvy-tripdata.csv', '202201-divvy-tripdata.csv', '202202-divvy-tripdata.csv', '202203-divvy-tripdata.csv', '202204-divvy-tripdata.csv', '202205-divvy-tripdata.csv')

  #files used are tallied
  number_of_files = length(file_list)
  
  #variable used in parsing loop
  number_of_loops = 1
  
  #loop for parsing data files
  #using this method means only 1 file at a time is kept in memory
  #This helps to make the system scaleable

  while (number_of_files >= number_of_loops) {
    # one month of data is read
    monthly_data <- read_csv(file_list[number_of_loops]) %>%
      
      #entries with no geolocation are dropped
      drop_na(start_lat, start_lng, end_lat, end_lng)
    
    # we shuffle the dataframe before we take a sample to ensure we are not creating a bias
    monthly_data = monthly_data[sample(1:nrow(monthly_data)), ] 
    
    # the number of data entries is counted and a number equal to 2% or 1/50th is stored
    # this number is used to reduce the data to a 2% sample size
    sample_size = as.integer(round(count(monthly_data)*.020, digits = 0)) 
    monthly_data = head(monthly_data, n = sample_size)
    
    #the month of data is used to create a file and consecutive months are added to the file
    if(number_of_loops == 1) 
      write_csv(monthly_data, file = "monthly_data_total.csv", append = TRUE, col_names = TRUE) 
    else   write_csv(monthly_data, file = "monthly_data_total.csv", append = TRUE, col_names = FALSE)
    
    #the loop repeats
    number_of_loops = number_of_loops + 1 }
  
```

### Data is loaded

Here is the file created by the code above.

```{r}
#dataframe is built from file
year_of_data <- read_csv('monthly_data_total.csv')
```

### Further Cleaning

-   Date values are converted and used in new columns.

-   Trip time is calculated and given a new column.

-   Trips with times less than one minute and greater than 1 day are removed.

-   A column is created that lists the day of the week, Monday through Sunday.

-   The new table is written to the dataframe.

```{r}
#this is the start of the "pipe" that will clean and alter our table
year_of_data %<>% 
  # null values in the longitude and latitude are removed
  drop_na(start_lat, start_lng, end_lat, end_lng) %>%
  #the date fields are changed into a different format
  mutate(started_at = as.POSIXct(started_at)) %>%
  mutate(ended_at = as.POSIXct(ended_at)) %>% 
  #columns for time and day are created
  mutate(ride_date = as.Date(started_at)) %>%
  mutate(ride_time = format(as.POSIXct(started_at, format='%Y-%m-%d %H:%M:%S'),format='%H:%M:%S')) %>%
  #a column for day of the week is created
  mutate(day_of_week = wday(started_at, label = TRUE, week_start = 1)) %>% 
  #ride time for each trip is calculated and returned as seconds
  mutate(ride_length = time_length(interval(started_at, ended_at))) %>%
  #trips shorter than 1 minute and longer than 1 day are dropped
  filter(ride_length > 60 & ride_length < 86400) %>% 
  #data is finalized and written to the dataframe
  select(rideable_type, member_casual, started_at, ended_at,
         ride_length, day_of_week, ride_date, ride_time, start_lat, start_lng, end_lat, end_lng, 
         start_station_name, end_station_name, start_station_id, end_station_id) 
```

## Visualizations are Created

**The following charts will illustrate how Cyclistic subscription members and casual (non_subscription) customers use the service differently.**

### Number of Trips On Each Weekday

```{r}
#bar chart for number of trips each weekday
ggplot(year_of_data, aes(x=day_of_week)) +
  geom_bar(aes(fill = member_casual), position = "dodge2") +
  scale_fill_manual(values=c("member" = "deepskyblue2", "casual" = "red")) +
  labs(title="Number of Trips on each Weekday", subtitle="Over one year", 
       y="Number of Trips", x="Weekday", fill="Rider type")
```

**The information above reveals that subscription members use the service more frequently during the week, while riders without a subscription use bikes more frequently on the weekend.**

### Daily Averages

```{r}
#a seperate data frame is created here to compute averages
mc_avg_ridetime_by_day <- year_of_data %>%
  group_by(ride_date, member_casual) %>%
  summarise_at(vars(ride_length), list(ride_time_avg = mean))

#a scatter plot is created showing how daily ride habits change over the year 
ggplot(mc_avg_ridetime_by_day, aes(x = ride_date, y = ride_time_avg)) +
  geom_point(aes(col=member_casual)) +
  geom_smooth(formula = y ~ x, span=.4, method = "loess", 
              aes(col=member_casual), show.legend = FALSE) +
  scale_color_manual(values = c("member" = "deepskyblue2", "casual" = "red")) +
  coord_cartesian(ylim=c(600, 6000), xlim=c(as.Date("2021-5-18"), as.Date("2022-5-14")))  +
  scale_x_continuous(breaks = seq(as.Date("2021-5-1"), as.Date("2022-6-1"), "1 month"),
    labels = scales::label_date("'%y %b"),
    limits = c(as.Date("2021-5-1"), as.Date("2022-5-31"))) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_y_continuous(breaks = seq(300, 6000, 900), label = c(" ", "15 min", "30 min", "45 min",
                                                             "1 hour", "1:15", "1:30")) +
  labs(title="Average Ridetime Each Day", subtitle="Each point is a daily average", 
       y="Average Time in Minutes", x="Months", col="Rider Type")
```

**The scatter-plot above shows us that casual customers ride for longer on average but trip times shorten during the winter months.**

### Daily Totals

```{r}
#a dataframe with the total number of rides by day is created
mc_trips_per_day <- year_of_data %>%
  group_by(ride_date = as.Date(ride_date), member_casual) %>%
  summarise(rides_per_day = n())
  
#rides per day chart
ggplot(mc_trips_per_day, aes(x=ride_date, y=rides_per_day)) + 
  geom_point(aes(col=member_casual)) +
  scale_color_manual(values = c("member" = "deepskyblue2", "casual" = "red")) +
  geom_smooth(formula = y ~ x, span=.4, method = "loess", 
              aes(col=member_casual), show.legend = FALSE) +
  coord_cartesian(xlim=c(as.Date("2021-5-18"), as.Date("2022-5-14")))  +
  scale_x_continuous(breaks = seq(as.Date("2021-5-1"), as.Date("2022-6-1"), "1 month"),
                     labels = scales::label_date("'%y %b"),
                     limits = c(as.Date("2021-5-1"), as.Date("2022-5-31"))) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
labs(title="Number of Uses per Day", subtitle="Each Point is a Daily Total", 
     y="Number of Rides per Day", x="Months", col="Rider Type")
```

**The above scatter-plot shows us the total number of rides per day. Notice the casual use customers' ridership peaks in August. We see a big drop-off in usage from casual customers during the winter months.**

## Conclusions

* The data would seem to indicate that subscription members use the service for a work commute.  
* Casual use customers are most active during the summer.
* Casual use customers ride for longer periods of time on average.

## Reccomendations

The best way to bring in more subscribers seems to be showing casual customers the benefits of using the service for a work commute.

A shorter subscription option may draw in customers who only want to bike during the warmer months of the year.

An even shorter option my be popular for visitors to the Chicago area.  One week or two week passes for people vacationing and on business trips may appeal to those who do not wish to rent a car.

Hot spots for marketing to casual users are listed below.  These are the station locations that casual users pick up bikes from most frequently.  A percentage was added to help illustrate how widely the service is used.  Most of these high use locations are in Central Chicago, with some high use in the north side and west side areas adjacent to central. 

```{r}
#table created showing frequent casual usage start locations
station_uses <- year_of_data %>%
  filter(member_casual == "casual") %>%
  drop_na(start_station_name) %>%  
  group_by(start_station_name) %>%
  summarize(total_station_uses = n()) %>%
  filter(total_station_uses > 50) %>%
  arrange(desc(total_station_uses)) %>%
  mutate(station_use_percent = sprintf("%.2f%%", 
                                       round((total_station_uses / sum(total_station_uses)*100), 
                                             digits = 2)))
#table created
knitr::kable(head(station_uses, 25), align = 'l')
```


